---
title: "STAT432 Assignment 2"
author: "Rory Sarten 301005654"
date: "21 August 2020"
output: pdf_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(magrittr)
library(ggplot2)
```

## Question 1

a) 

These are independent binomial processes (despite happening in sequence), so the three trapping occasions can be denoted:

$$n_1 \sim Binomial(N, \theta) = \binom{N}{n_{1}} \theta^{n_1} (1 - \theta)^{N - n_1}$$
$$n_2 \sim Binomial(N - n_1, \theta) = 
\binom{N - n_1}{n_{2}} \theta^{n_2} (1 - \theta)^{N - n_1 - n_2}$$

$$n_3 \sim Binomial(N - n_1 - n_2, \theta) = 
\binom{N - n_1 - n_2}{n_{3}} \theta^{n_3} (1 - \theta)^{N - n_1 - n_2 - n_3}$$

So the probability of catching $n_1 + n_2 + n_3$ rats is given by 

$$Binomial(N, \theta) \times Binomial(N - n_1, \theta) \times Binomial(N - n_1 - N-2, \theta)$$

b)

$$\ell(N,\theta)=\ln \left[ \dfrac{N!}{n_1(N-n_1)!}\theta^{n_1}(1-\theta)^{N-n_1}\times\dfrac{(N-n_1)!}{n_2(N-n_1-n_2)!}\theta^{n_2}(1-\theta)^{N-n_1-n_2}\times\dfrac{(N-n_1-n_2)!}{n_3(N-n_1-n_2-n_3)!}\theta^{n_3}(1-\theta)^{N-n_1-n_2-n_3}\right]$$

$$\ell(N,\theta) = 
\ln N! - [\ln n_1 + \ln (N-n_1)] + n_1 \ln \theta + (N-n_1)\ln (1 - \theta) +$$
$$\ln (N-n_1)! - [\ln n_2 + \ln (N-n_1-n_2)] + n_2 \ln \theta + (N-n_1-n_2)\ln (1 - \theta) +$$
$$\ln (N-n_1-n_2)! - [\ln n_3 + \ln (N-n_1-n_2-n_3)] + n_3 \ln \theta + (N-n_1-n_2-n_3)\ln (1 - \theta)$$

finish that bit of algebra somehow

c)

```{r solve_c}
llfunc <- function(par, n1, n2, n3) {
  N <- par[1]
  theta <- par[2]
  lfactorial(N) - lfactorial(N - n1 - n2 - n3) +
    sum(n1, n2, n3)*log(theta) + 
    (3*N - 3*n1 - 2*n2 - n3)*log(1 - theta)
}

n1 <- 82L
n2 <- 54L
n3 <- 23L
## since N >= n1 + n2 + n3
N_0 <- n1 + n2 + n3
par_start <- c(N_0, 0.5)

optim_fit <- optim(par = par_start,
                   fn = llfunc,
                   n1 = n1, n2 = n2, n3 = n3,
                   lower = c(N_0, 1e-4),
                   upper = c(Inf, 1),
                   method = "L-BFGS-B",
                   control = list(fnscale = -1),
                   hessian = TRUE)

MLE <- optim_fit$par
SE <- sqrt(diag(solve(-optim_fit$hessian)))
LowerBound <- MLE - qnorm(0.975) * SE
UpperBound <- MLE + qnorm(0.975) * SE
results <- cbind(MLE, SE, LowerBound, UpperBound) %>% round(2)
rownames(results) <- c("N", "theta")
results
```

The maximum likelihood point $\hat{N} = 189.43$ with confidence interval $(163.49, 215.36)$

The maximum likelihood point $\hat{\theta} = 0.45$ with confidence interval $(0.34, 0.57)$

d)

Contour map of $(N, \theta)$ surface

```{r contour_map}
n_vals <- seq(from = n1+n2+n3, to = 260, length = 100)
theta_vals <- seq(from = 0.2, to = 0.7, length = 100)

combos <- expand.grid(n_vals, theta_vals)
surface <- apply(expand.grid(n_vals, theta_vals), 1, llfunc, n1 = n1, n2 = n2, n3 = n3) %>% round(2)
outcome <- cbind(combos, surface)
names(outcome) <- c("N", "theta", "value")

ggplot(data = outcome, mapping = aes(N, theta)) + 
  geom_contour(aes(z = value),
                   breaks = seq(from = min(outcome$value), to = max(outcome$value), by = 12)) +
  geom_contour(aes(z = value,
                   colour = factor(..level.. == max(surface) - 1.92,
                                   levels = c(F, T),
                                   labels = c("Single 95% CI "))),
               breaks = max(surface) - 1.92) +
  geom_contour(aes(z = value,
                   colour = factor(..level.. == max(surface) - 3,
                                   levels = c(F, T),
                                   labels = c("Joint 95% CI "))),
               breaks = max(surface) - 3) +
  geom_point(mapping = aes(x = MLE[1], y = MLE[2]), colour = "darkblue") +
  labs(colour = "Of interest:") +
  #ggtitle("Contour map of (N, \u03B8) surface") +
  #ylab("\u03B8") +
  theme_minimal()
```



things